


For the differential equation:

$$\dfrac{dy}{dt} = y + 2t$$
the general analytical solution is:
$$y(t) = C_1e^{t} - 2t - 2$$

Find the errors given this initial condition: 
$$y(0) = 1$$
which allows to find:
$$C_1 = 3$$


```{r message=FALSE, results="hold"}
library(rODE)

setClass("MuskatTest", slots = c(
    stack = "environment"           # environment object inside the class
    ),
    contains = c("ODE")
    )

setMethod("initialize", "MuskatTest", function(.Object, ...) {
    .Object@stack$n <-  0               # "n" belongs to the class environment
    .Object@state   <- c(1.0, 0.0)      # initial value
    return(.Object)
})

setMethod("getExactSolution", "MuskatTest", function(object, t, ...) {
    # analytical solution
    return(3 * exp(t) - 2 *t - 2)       # constant C1 = 3 
})

setMethod("getState", "MuskatTest", function(object, ...) {
    object@state
})

setMethod("getRate", "MuskatTest", function(object, state, ...) {
    object@rate[1] <- state[1] + 2 * state[2]
    object@rate[2] <-  1                        # rate of change of time, dt/dt
    object@stack$n <-  object@stack$n + 1       # add 1 to the rate count
    object@rate
})

# constructor
MuskatTest <- function() {
    odetest <- new("MuskatTest")
    odetest
}
```

## Euler

```{r}
ComparisonEulerApp <- function(stepSize) {
    ode <- new("MuskatTest")
    ode_solver <- Euler(ode)
    ode_solver <- setStepSize(ode_solver, stepSize)
    time <-  0
    rowVector <- vector("list")
    i <-  1
    while (time < 5) {
        state <- getState(ode_solver@ode)
        time <- state[2]
        error <- getExactSolution(ode_solver@ode, time) - state[1]
        rowVector[[i]] <- list(step_size = stepSize, 
                               t = time, 
                               y_t = state[1], 
                               exact = getExactSolution(ode_solver@ode, time),
                               error = error, 
                               rel_err = error / getExactSolution(ode_solver@ode, time),
                               steps = ode_solver@ode@stack$n
                               )
        ode_solver <- step(ode_solver)
        i <- i + 1
    }
    data.table::rbindlist(rowVector)
}
```

```{r}
# get a summary table for different step sizes
get_table <- function(stepSize) {
    dt <- ComparisonEulerApp(stepSize)
    dt[round(t,1) %in% c(1.0, 2.0, 3.0, 4.0, 5.0)]
}

step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, get_table)
data.table::rbindlist(dt_li)
```

## Using RK4

```{r}
ComparisonRK4App <- function(stepSize) {
    ode <- new("MuskatTest")
    ode_solver <- RK4(ode)
    ode_solver <- setStepSize(ode_solver, stepSize)
    time <-  0
    rowVector <- vector("list")
    i <-  1
    while (time < 5) {
        state <- getState(ode_solver@ode)
        time  <- state[2]
        error <- getExactSolution(ode_solver@ode, time) - state[1]
        rowVector[[i]] <- list(step_size = stepSize, 
                               t = time, 
                               y_t = state[1], 
                               exact = getExactSolution(ode_solver@ode, time),
                               error = error, 
                               rel_err = error / getExactSolution(ode_solver@ode, time),
                               steps = ode_solver@ode@stack$n
                               )
        ode_solver <- step(ode_solver)
        i <- i + 1
    }
    data.table::rbindlist(rowVector)
}

# get a summary table for different step sizes
get_table <- function(stepSize) {
    dt <- ComparisonRK4App(stepSize)
    dt[round(t, 1) %in% c(1.0, 2.0, 3.0, 4.0, 5.0)]
}

step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, get_table)
data.table::rbindlist(dt_li)
```

We see above that we are repeating code when the only parameter that is being changed is the ODE solver. In these cases is more convenient to use the function `ODESolverFactory`.


## Using a solver factory

```{r}
ComparisonApp <- function(solver, stepSize) {
    ode <- new("MuskatTest")
    solver_factory <- ODESolverFactory()
    ode_solver <- createODESolver(solver_factory, ode, solver)
    ode_solver <- setStepSize(ode_solver, stepSize)
    rowVector <- vector("list")
    time <-  0
    i <-  1
    while (time < 5.001) {
        state <- getState(ode_solver@ode)
        time  <- state[2]
        error <- getExactSolution(ode_solver@ode, time) - state[1]
        rowVector[[i]] <- list(step_size = stepSize, 
                               t = time, 
                               y_t = state[1], 
                               exact = getExactSolution(ode_solver@ode, time),
                               error = error, 
                               rel_err = error / getExactSolution(ode_solver@ode, time),
                               steps = ode_solver@ode@stack$n
                               )
        ode_solver <- step(ode_solver)
        time <- time + getStepSize(ode_solver)    # step size retrievd from ODE solver
        i <- i + 1
    }
    data.table::rbindlist(rowVector)
}

# get a summary table for different step sizes
create_table <- function(stepSize, solver) {
    dt <- ComparisonApp(solver, stepSize)
    # dt
    dt[round(t, 1) %in% c(1.0, 2.0, 3.0, 4.0, 5.0)]
}
```

### Euler
```{r}
# Create summary table for ODE solver Euler
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, create_table, solver = "Euler")
data.table::rbindlist(dt_li)
```

### Verlet

```{r}
# Create summary table for ODE solver Verlet
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, create_table, solver = "Verlet")
data.table::rbindlist(dt_li)
```

### Euler-Richardson
```{r}
# Create summary table for ODE solver EulerRichardson
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, create_table, solver = "EulerRichardson")
data.table::rbindlist(dt_li)
```

### Runge-Kutta

```{r}
# Create summary table for ODE solver RK4
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, create_table, solver = "RK4")
data.table::rbindlist(dt_li)
```

### Runge-Kutta 45

```{r}
# Create summary table for ODE solver RK45
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, create_table, solver = "RK45")
data.table::rbindlist(dt_li)
```


