---
title: "Muskat Material Balance"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Muskat Material Balance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

For the Muskat's Material Balance equation:

$$\newcommand{\numD}{{\dfrac {S_o}{B_o B_g} \dfrac {dR_s}{dP} +  
\dfrac {S_o}{B_o} \dfrac {k_g}{k_o} \dfrac {\mu_o}{\mu_g} \dfrac {dB_o}{dP} + 
(1 - S_o - S_w) \dfrac {1}{B_g} \dfrac {dB_g}{dP} }}$$


$$\dfrac {dS_o}{dP} = \dfrac {\numD} {1 + \dfrac {k_g}{k_o} \dfrac {\mu_o}{\mu_g} }$$

All the terms on the right side are function of pressure ($P$) and saturation ($S$), the equation could be reduced to:

$$ \dfrac{dS}{dP} = f(P, S)$$

A first-order ordinary diferential equation (*ODE*).

With analytical solution:

$$y(x) = C_1e^x-2x-2$$

## Plot a range of values for a given step size

Given the ODE:
$$ \dfrac{dS}{dP} = (2P +  S)$$

Find the first saturation values if the initial conditions are $P_o$ = 0, and $S_o$ = 1.

The analytical solution for the the differential equation given the initial conditions is:
$$S(P) = 3e^P-2P-2$$


## Build the ODE object
Since we know the analytical solution, we will add a method `getExactSolution` to return the exact values at any given P.

```{r message=FALSE, results="hold"}
# the ODE object

library(rODE)
library(ggplot2)

setClass("MuskatODE", slots = c(
    stack = "environment"                # environment object inside the class
    ),
    contains = c("ODE")
    )

setMethod("initialize", "MuskatODE", function(.Object, ...) {
    .Object@stack$n <-  0               
    return(.Object)
})

setMethod("getExactSolution", "MuskatODE", function(object, t, ...) {
    # analytical solution
    return(3 * exp(t) - 2 *t - 2)       # constant C1 = 3 
})

setMethod("getState", "MuskatODE", function(object, ...) {
    object@state
})

setMethod("getRate", "MuskatODE", function(object, state, ...) {
    object@rate[1] <- state[1] + 2 * state[2]         # 2P + S
    object@rate[2] <- 1                               # dP/dP
    object@stack$n <- object@stack$n + 1              # add 1 to the rate count
    object@rate
})

# constructor
MuskatODE <- function(P, S) {
    .MuskatEuler <- new("MuskatODE")
    .MuskatEuler@state[1] = S        # S
    .MuskatEuler@state[2] = P        # P = t
    return(.MuskatEuler)
}
```

## Implementing the Euler solver

```{r}
MuskatEulerApp <- function(stepSize) {
    ode <- MuskatODE(0, 1)               # initial state  S(0) = 1
    ode_solver <- Euler(ode)
    ode_solver <- setStepSize(ode_solver, stepSize)
    rowVector <- vector("list")
    pres <-  0
    i    <-  1
    while (pres < 5) {
        state <- getState(ode_solver@ode)
        pres <- state[2]
        error <- getExactSolution(ode_solver@ode, pres) - state[1]
        rowVector[[i]] <- list(step_size = stepSize, 
                               P = pres, 
                               S = state[1], 
                               exact = getExactSolution(ode_solver@ode, pres),
                               error = error, 
                               rel_err = error / getExactSolution(ode_solver@ode, pres),
                               steps = ode_solver@ode@stack$n
                               )
        ode_solver <- step(ode_solver)
        i <- i + 1
    }
    data.table::rbindlist(rowVector)
}
```

### Summary table for Euler

```{r}
# get a summary table for different step sizes
get_table <- function(stepSize) {
    dt <- MuskatEulerApp(stepSize)
    dt[round(P, 1) %in% c(1.0, 2.0, 3.0, 4.0, 5.0)]   # round at 1 decimal
}
# vector with some step sizes
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, get_table)    # create a data table 
data.table::rbindlist(dt_li)              # bind the data tables
```


## Solve Muskat's Equation using Runge-Kutta

```{r}
MuskatRK4App <- function(stepSize) {
    ode <- MuskatODE(0, 1)
    ode_solver <- RK4(ode)
    ode_solver <- setStepSize(ode_solver, stepSize)
    rowVector <- vector("list")
    pres <-  0
    i    <-  1
    while (pres < 5) {
        state <- getState(ode_solver@ode)
        pres <- state[2]
        error <- getExactSolution(ode_solver@ode, pres) - state[1]
        rowVector[[i]] <- list(step_size = stepSize, 
                               P = pres, 
                               S = state[1], 
                               exact = getExactSolution(ode_solver@ode, pres),
                               error = error, 
                               rel_err = error / getExactSolution(ode_solver@ode, pres),
                               steps = ode_solver@ode@stack$n
                               )
        ode_solver <- step(ode_solver)
        i <- i + 1
    }
    data.table::rbindlist(rowVector)
}

# get a summary table for different step sizes
get_table <- function(stepSize) {
    dt <- MuskatRK4App(stepSize)
    dt[round(P, 1) %in% c(1.0, 2.0, 3.0, 4.0, 5.0)]
}

step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, get_table)
data.table::rbindlist(dt_li)
```

We see above that we are repeating code when the only parameter needed to be changed is the ODE solver. In cases where we want to test different ODE solvers it is more convenient to use the function `ODESolverFactory` and send the solver as a parameter.


## Using a solver factory

```{r}
ComparisonMuskatODEApp <- function(solver, stepSize) {
    ode <- MuskatODE(0, 1)
    solver_factory <- ODESolverFactory()
    ode_solver <- createODESolver(solver_factory, ode, solver)
    ode_solver <- setStepSize(ode_solver, stepSize)
    rowVector  <- vector("list")
    pres <-  0
    i    <-  1
    while (pres < 5.001) {
        state <- getState(ode_solver@ode)
        pres  <- state[2]
        error <- getExactSolution(ode_solver@ode, pres) - state[1]
        rowVector[[i]] <- list(solver = solver,
                               step_size = stepSize, 
                               P = pres, 
                               S = state[1], 
                               exact = getExactSolution(ode_solver@ode, pres),
                               error = error, 
                               rel_err = error / getExactSolution(ode_solver@ode, pres),
                               steps = ode_solver@ode@stack$n
                               )
        ode_solver <- step(ode_solver)
        pres <- pres + getStepSize(ode_solver)    # step size retrievd from ODE solver
        i <- i + 1
    }
    data.table::rbindlist(rowVector)
}

# get a summary table for different step sizes
create_table <- function(stepSize, solver) {
    dt <- ComparisonMuskatODEApp(solver, stepSize)
    # dt
    dt[round(P, 1) %in% c(1.0, 2.0, 3.0, 4.0, 5.0)]
}
```


### Euler
```{r}
# Create summary table for ODE solver Euler
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, create_table, solver = "Euler")
data.table::rbindlist(dt_li)
```

### Euler-Richardson
```{r}
# Create summary table for ODE solver EulerRichardson
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, create_table, solver = "EulerRichardson")
data.table::rbindlist(dt_li)
```

### Runge-Kutta

```{r}
# Create summary table for ODE solver RK4
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, create_table, solver = "RK4")
data.table::rbindlist(dt_li)
```

### Runge-Kutta 45

```{r}
# Create summary table for ODE solver RK45
step_sizes <- c(0.2, 0.1, 0.05)
dt_li <- lapply(step_sizes, create_table, solver = "RK45")
data.table::rbindlist(dt_li)
```

And so on. We could this better. We will show how to do this using nested `lapply` functions.

## What about doing all in one step

```{r}
# vectors for the solvers and step sizes
step_sizes <- c(0.2, 0.1, 0.05)
solvers <- c("Euler", "EulerRichardson", "Verlet", "RK4", "RK45")
```


```{r}
# nested lapply to iterate through solvers and step sizes
df_li <- lapply(solvers, function(svr)
            lapply(step_sizes, function(stepsz) create_table(stepsz, svr)))

# join the resulting dataframes
df_all <- data.table::rbindlist(unlist(df_li, recursive = FALSE))
df_all
```

### Plot the error of the solvers

```{r}
ggplot(df_all, aes(x = P, y = rel_err, group = step_size, fill = solver )) +
    geom_line() + 
    geom_area(stat = "identity") + 
    facet_grid(step_size ~ solver)
```

In this last plot we want to compare the relative error of the ODE solvers that show the least error: RK4 and RK45. At the same time, we will exclude the step size of 0.2 since its error magnitude would hide those of the smaller steps.

### Plot RK4 vs RK45

```{r}
ggplot(subset(df_all, solver %in% c("RK4", "RK45") & step_size %in% c(0.1, 0.05)), aes(x = P, y = rel_err, group = step_size, fill = solver )) +
    geom_line() + 
    geom_area(stat = "identity") + 
    facet_grid(step_size ~ solver) 
```






